---
title: "Initial exploration"
author: "Martine Wauben"
date: "01/05/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.width = 12)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(zoo)

strong_palette <- c("#00aae6", "#ff8400", "#00d1fc")
soft_palette <- c("#dcf4fe", "#fff0df")

BIT_theme <- theme_minimal()+
  theme(plot.title = element_text(color = "#00aae6",
                                  size = rel(1.8)),
        plot.subtitle = element_text(color = "#00aae6"),
        strip.text.x = element_text(size = rel(1.1), 
                                    colour = "black",
                                    margin = margin(t = 2,
                                                    b = 2)),
        strip.background = element_rect(fill = "#dcf4fe", colour = NA))

```


# Data preparation

## Uni / Bivariate visualisations

Load data, calculate variables required, and calculate a dataset which allows you to calculate trend lines over time. That is, I infer which individuals were in the workforce in any one year based on their hire date and tenure.

```{r data prep}
# Loading and calculating data
data <- read.csv('GCR.csv') %>%
  dplyr::mutate(gender = factor(gender, levels = c("M", "F")),
                gradeAtHire = as.ordered(gradeAtHire),
         age = ageAtHire + tenure,
         gend_dept = paste(gender, department, sep = "_"),
         leaveYear = hireYear + tenure,
         term = ifelse(reason == "terminated", 1, 0))
# str(data)
summary(data)

# Workforce over time: ----
all_years <- expand.grid(year = unique(c(data$hireYear, data$leaveYear)),
                         id = unique(data$id))
data_timeline <- dplyr::full_join(data, all_years, by = "id") %>%
  dplyr::filter(hireYear <= year & leaveYear >= year) %>%
  dplyr::group_by(id) %>%
  dplyr::mutate(reason = ifelse(year != max(year), NA, as.character(reason)),
                current = ifelse(year != max(year), TRUE, FALSE),
                term = ifelse(year != max(year), NA, as.numeric(term)),
                new = hireYear == year)

# Proportion of women in the workforce any year: ----
workforce_prop_timeline <- data_timeline %>%
  dplyr::group_by(year, gender) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(prop_workforce = n / sum(n)) %>%
  dplyr::filter(gender == "F") 


# People who have left: ----
data_left <- data[!data$current, ]
```


We assume that the business was founded in `r min(data$hireYear)`. 

# Data exploration

I explored univariate and bivariate distributions of variables I hypothesise may be related to each other and to gender using some basic visualisations. If you are interested, click the buttons below.

<p>
  <a class="btn btn-primary" data-toggle="collapse" href="#collapse1" role="button" aria-expanded="false" aria-controls="collapse1">
    Click for visualisations of the current workforce
  </a>
</p>
<div class="collapse" id="collapse1">
  <div class="card card-body">

```{r visualise current workforce, message=FALSE}
data_current <- data[data$current, ]
summary(data_current)
# Age by gender
ggplot(data_current, aes(x = gender, y = age))+
  geom_violin()+
  BIT_theme
# Tenure by gender
ggplot(data_current, aes(x = gender, y = tenure))+
  geom_violin()+
  BIT_theme
# Department by gender
ggplot(data_current, aes(x = gender, y = department))+
  geom_bin2d()+
  BIT_theme
# Grade at hire by gender
ggplot(data_current, aes(x = gender, y = gradeAtHire))+
  geom_bin2d()+
  BIT_theme
# Tenure by department
ggplot(data_current, aes(x = department, y = tenure))+
  geom_boxplot()+
  BIT_theme
# Age by department
ggplot(data_current, aes(x = department, y = age))+
  geom_violin()+
  BIT_theme

# Age by gender and department
ggplot(data_current, aes(x = gend_dept, y = tenure))+
  geom_violin()+
  BIT_theme
# Mainly the administrative dept has much less experienced women than men, much more similar in operations
ggplot(data_current, aes(x = gend_dept, y = age))+
  geom_violin()+
  BIT_theme
ggplot(data_current, aes(x = gend_dept, y = ageAtHire))+
  geom_violin()+
  BIT_theme
```

</div>
</div>


<p>
  <a class="btn btn-primary" data-toggle="collapse" href="#collapse2" role="button" aria-expanded="false" aria-controls="collapse2">
    Click for visualisations of the previous leavers
  </a>
</p>
<div class="collapse" id="collapse2">
  <div class="card card-body">

```{r visualise leavers, message=FALSE}
# Were there peak leaving years?
hist(data_left$leaveYear, breaks = length(unique(data_left$leaveYear)))
# There were some spikes: 1986, 2002, 2008/9
# Were women overrepresented in those years?
data_years <- data_left %>%
  dplyr::group_by(leaveYear, gender) %>%
  dplyr::summarise(n = n()) %>%
  tidyr::spread(gender, n) %>%
  dplyr::mutate(F_prop = `F` / (`F` + `M`))
ggplot(data_years, aes(x = leaveYear, y = F_prop))+
  geom_bar(stat = "identity")
# Not noticeably more women leaving in those years as a proportion of everyone leaving that year

# Age by gender
ggplot(data_left, aes(x = gender, y = age))+
  geom_violin()+
  BIT_theme
# Tenure by gender
ggplot(data_left, aes(x = gender, y = tenure))+
  geom_violin()+
  BIT_theme
# Department by gender
ggplot(data_left, aes(x = gender, y = department))+
  geom_bin2d()+
  BIT_theme
# Grade at hire by gender
ggplot(data_left, aes(x = gender, y = gradeAtHire))+
  geom_bin2d()+
  BIT_theme
# Tenure by department
ggplot(data_left, aes(x = department, y = tenure))+
  geom_boxplot()+
  BIT_theme
# Age by department
ggplot(data_left, aes(x = department, y = age))+
  geom_violin()+
  BIT_theme

# Age by gender and department
ggplot(data_left, aes(x = gend_dept, y = tenure))+
  geom_violin()+
  BIT_theme
# Mainly the administrative dept has much less experienced women than men, much more similar in operations
ggplot(data_left, aes(x = gend_dept, y = age))+
  geom_violin()+
  BIT_theme
ggplot(data_left, aes(x = gend_dept, y = ageAtHire))+
  geom_violin()+
  BIT_theme
```

</div>
</div>

## Workforce over time

Calculate the proportion of women in each section of the workforce (department) in a given year, and compare that to the overall proportion of women in the whole workforce.

```{r workforce in any year, message=FALSE}
data_dept_timeline <- data_timeline  %>%
  dplyr::group_by(year, department, gender) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::filter(gender == "F") %>%
  left_join(., workforce_prop_timeline[, c("year", "prop_workforce")], by = "year")
ggplot(data_dept_timeline, aes(x = year, y = prop, colour = department))+
  geom_line()+
  geom_smooth(alpha = 0.25)+
  geom_line(aes(x = year, y = prop_workforce), 
              colour = "red")+
  labs(title = "Proportion of female workforce over time",
       x = "Year",
       y = "Proportion of workers that were female",
       colour = "Department",
       subtitle = "GCR overall proportion in shown in red")+
  BIT_theme+
  scale_colour_manual(values = strong_palette)
```

## Reasons for leaving

Starting from some basic density plots, I can calculate the proportion of leavers who are women for each separate leaving reason. This tells me whether there are specific reasons that women leave for more often than the general population. Because I saw above that there is a significant difference between the numbers of women between each department, I calculate the reasons separately for each department. 

```{r Reasons for leaving, message=FALSE}
# Reasons for leaving by gender and department
ggplot(data_left, aes(x = gend_dept, y = reason))+
  geom_bin2d()+
  BIT_theme+
  scale_fill_gradientn(colors=strong_palette[1:2])

workforce_dept_prop_timeline <- data_timeline %>%
  dplyr::group_by(year, department, gender) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::mutate(prop_workforce = n / sum(n)) %>%
  dplyr::filter(gender == "F") %>%
  dplyr::select(leaveYear = year, department, prop_workforce)

all_reasons <- expand.grid(leaveYear = unique(data_left$leaveYear),
                           department = unique(data_left$department),
                           reason = unique(data_left$reason),
                           gender = unique(data_left$gender))
data_reasons <- data_left %>%
  dplyr::group_by(leaveYear, department, reason, gender) %>%
  dplyr::summarise(n = n()) %>%
  dplyr::ungroup() %>%
  dplyr::full_join(., all_reasons, 
                   by = c("leaveYear", "department", "reason", "gender")) %>%
  dplyr::mutate(n = ifelse(is.na(n), 0, n)) %>%
  dplyr::group_by(leaveYear, department, reason) %>%
  dplyr::mutate(tot = sum(n)) %>%
  dplyr::ungroup() %>%
  dplyr::filter(gender == "F") %>%
  dplyr::mutate(prop = ifelse(tot > 0,
                              n / tot, 0)) %>%
  dplyr::left_join(., workforce_dept_prop_timeline, 
                   by = c("leaveYear", "department")) %>%
  dplyr::arrange(leaveYear, department, reason, gender)


ggplot(data_reasons, aes(x = leaveYear, y = prop, colour = reason))+
  geom_line(alpha = 0.5)+
  geom_smooth(se = FALSE)+
  geom_line(aes(x = leaveYear, y = prop_workforce), 
            colour = "red")+
  facet_wrap(department~reason)+
  labs(title = "Women as a % of all who leave for each reason",
       x = "Year of leaving",
       y = "Proportion of women",
       subtitle = "Department's overall proportion of women shown in red")+
  BIT_theme+
  scale_colour_manual(values = strong_palette)

# Significant association between gender and being terminated?
reason_assoc <- chisq.test(table(data$gender, data$term))
```

There appears to be a strong association between gender and being terminated, at a Chi squared of `r round(reason_assoc$statistic, 2)`, p `r ifelse(round(reason_assoc$p.value, 3) > 0, paste0("= ", round(reason_assoc$p.value, 3)), "< .001")`.

# Data analysis

I want to dig deeper into this association between gender and being terminated. First, I created a set of mosaic plots to explore the bivariate association between termination and other variables. Click the button below to see those. 

<p>
  <a class="btn btn-primary" data-toggle="collapse" href="#collapse3" role="button" aria-expanded="false" aria-controls="collapse3">
    Click for visualisations bivariate associations with termination
  </a>
</p>
<div class="collapse" id="collapse3">
  <div class="card card-body">

```{r characteristics of terminated people, message=FALSE}
library("graphics")
data_dept <- as.table(as.matrix(table(data_left$department, data_left$term)))
mosaicplot(data_dept, shade = TRUE, las=2, main = "Department")
data_ageH <- as.table(as.matrix(table(data_left$ageAtHire, data_left$term)))
mosaicplot(data_ageH, shade = TRUE, las=2, main = "ageAtHire")
data_ageL <- as.table(as.matrix(table(data_left$age, data_left$term)))
mosaicplot(data_ageL, shade = TRUE, las=2, main = "ageWhenLeft")
data_year <- as.table(as.matrix(table(data_left$hireYear, data_left$term)))
mosaicplot(data_year, shade = TRUE, las=2, main = "hireYear")
data_leave <- as.table(as.matrix(table(data_left$leaveYear, data_left$term)))
mosaicplot(data_leave, shade = TRUE, las=2, main = "leaveYear")
# data_ten <- as.table(as.matrix(data_left$tenure, data_left$term))
# mosaicplot(data_ten, shade = TRUE, las=2, main = "Tenure")
data_gend <- as.table(as.matrix(table(data_left$gender, data_left$term)))
mosaicplot(data_gend, shade = TRUE, las=2, main = "Gender")
data_grade <- as.table(as.matrix(table(data_left$gradeAtHire, data_left$term)))
mosaicplot(data_grade, shade = TRUE, las=2, main = "GradeAtHire")
```

</div>
</div>

I then formalised these analyses with an ANOVA analysis on termination.

```{r anova on termination, message=FALSE}
summary(aov(term ~ gradeAtHire * department * gender * ageAtHire * leaveYear, data = data_left))
# Main effects of department, gender, and hireYear, but 3-way interaction of GradeAtHire, gender and HireYear
```

And visualised the three-way interaction I observed in the analysis.

```{r vis termination interaction, message=FALSE}
data_term_leave <- data_left %>%
  dplyr::group_by(leaveYear, department, gender) %>%
  dplyr::summarise(leave_left = n(),
                   leave_terminated = sum(term),
                   leave_prop_term = sum(term) / n(),
                   m_ageAtHire = mean(ageAtHire),
                   sd_ageAtHire = sd(ageAtHire, na.rm = T),
                   m_ageWhenLeft = mean(age),
                   sd_ageWhenLeft = sd(age),
                   m_tenure = mean(tenure),
                   sd_tenure = sd(tenure),
                   m_gradeAtHire = mean(as.numeric(gradeAtHire)),
                   sd_gradeAtHire = sd(as.numeric(gradeAtHire))) %>%
  dplyr::select("year" = leaveYear, department, gender, leave_left, leave_terminated, leave_prop_term)


r1 <- ggplot(data_term_leave, aes(x = year, y = leave_prop_term, colour = gender))+
  geom_line(alpha = 0.5)+
  geom_smooth(se = FALSE)+
  facet_wrap(~department)+
  labs(title = "Of those who left, the proportion terminated",
       x = "Year left",
       y = "Proportion terminated",
       colour = "Gender",
       subtitle = "By department, gender and year they left")+
  BIT_theme+
  scale_colour_manual(values = strong_palette)

data_term_hire <- data_left %>%
  dplyr::group_by(hireYear, department, gender) %>%
  dplyr::summarise(hire_left = n(),
                   hire_terminated = sum(term),
                   hire_prop_term = sum(term) / n(),
                   m_ageAtHire = mean(ageAtHire),
                   sd_ageAtHire = sd(ageAtHire, na.rm = T),
                   m_ageWhenLeft = mean(age),
                   sd_ageWhenLeft = sd(age),
                   m_tenure = mean(tenure),
                   sd_tenure = sd(tenure),
                   m_gradeAtHire = mean(as.numeric(gradeAtHire)),
                   sd_gradeAtHire = sd(as.numeric(gradeAtHire))) %>%
  select("year" = hireYear, department, gender, hire_left, hire_terminated, hire_prop_term)


r2 <- ggplot(data_term_hire, aes(x = year, y = hire_prop_term, colour = gender))+
  geom_line(alpha = 0.5)+
  geom_smooth(se = FALSE)+
  facet_wrap(~department)+
  labs(title = "Of those who left, the proportion terminated",
       x = "Year hired",
       y = "Proportion terminated",
     subtitle = "By department, gender and year they were hired")+
  BIT_theme+
  scale_colour_manual(values = strong_palette)

grid.arrange(r1, r2, ncol = 1)
```



```{r rate of termination for men and women, message=FALSE}
data_rate <- data_timeline %>%
  dplyr::group_by(year, department, gender) %>%
  dplyr::summarise(workforce = n(),
                   left = sum(current != TRUE),
                   joined = sum(new == TRUE),
                   terminated = sum(term, na.rm = T)) %>%
  dplyr::mutate(joining_rate = joined / workforce,
                leaving_rate = left / workforce,
                termination_rate = terminated / workforce,
                term_leaving_rate = terminated / left) %>%
  dplyr::select(year, department, gender, 
                "Joining rate" = joining_rate, 
                "Leaving rate" = leaving_rate, 
                "Termination rate - workforce" = termination_rate, 
                "Terminations - within leavers" = term_leaving_rate) %>%
  tidyr::gather("rate", "value", -year, -department, -gender)

ggplot(data_rate, aes(x = year, y = value, colour = gender))+
  geom_line(alpha = 0.5)+
  geom_smooth(se = FALSE)+
  facet_wrap(department~rate, ncol = 4, scales = "free")+
  labs(title = "Rates of joining, leaving and termination in the workforce",
       x = "Year",
       y = "Rate",
       colour = "Gender",
       subtitle = "As a proportion of the male or female workforce respectively, or of leavers only")+
  BIT_theme+
  scale_colour_manual(values = strong_palette)
```

# Male and female firing rates

So what would have happened if women had been fired at the same rate as men for the last few years?

First, I calculated a 5-year rolling average of the rate of termination for men, and used that to calculate the female workforce based on the actual workforce.

```{r termination rate, message=FALSE}

# Workforce forecast: -----
yeargenderreason <- expand.grid(year = unique(data_timeline$year),
                        gender = unique(data_timeline$gender),
                        reason = unique(data_timeline$reason))
new_year <- data_timeline %>%
  dplyr::group_by(year, gender) %>%
  summarise(new = sum(new))
workforce_trend <- data_timeline %>%
  dplyr::group_by(year, gender, reason) %>%
  dplyr::summarise(number = n()) %>%
  merge(., yeargenderreason, all.x = T, all.y = T, by = c("year", "gender", "reason")) %>%
  dplyr::mutate(number = ifelse(is.na(number), 0, number),
                reason = ifelse(is.na(reason), "stayed", reason)) %>%
  tidyr::spread(reason, number) %>%
  merge(., new_year, all.x = T, all.y = T, by = c("year", "gender")) %>%
  dplyr::mutate(total = stayed + retired + resigned + terminated) %>%
  dplyr::mutate(join_rate = new / total,
                term_rate = terminated / total,
                nonterm_rate = (retired + resigned) / total) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(gender) %>%
  dplyr::mutate(join_avg = zoo::rollmean(join_rate, 5, na.pad = T, align = "right"),
                term_avg = zoo::rollmean(term_rate, 5, na.pad = T, align = "right"),
                nonterm_avg = zoo::rollmean(nonterm_rate, 5, na.pad = T, align = "right"))

f_rates <- workforce_trend %>%
  dplyr::filter(year == 2017) %>%
  dplyr::mutate(type = "f_rates",
                year1 = total + total * join_avg - total * term_avg - total * nonterm_avg) %>%
  dplyr::mutate(year2 = year1 + year1 * join_avg - year1 * term_avg - year1 * nonterm_avg) %>%
  dplyr::mutate(year3 = year2 + year2 * join_avg - year2 * term_avg - year2 * nonterm_avg) %>%
  dplyr::mutate(year4 = year3 + year3 * join_avg - year3 * term_avg - year3 * nonterm_avg) %>%
  dplyr::mutate(year5 = year4 + year4 * join_avg - year4 * term_avg - year4 * nonterm_avg) %>%
  dplyr::select(gender, type, `2017` = total, `2018` = year1, 
                `2019` = year2, `2020` = year3, `2021` = year4, `2022` = year5)
m_join_avg <- workforce_trend %>%
  dplyr::filter(gender == "M", year == 2017) %>%
  .$join_avg
m_term_avg <- workforce_trend%>%
  dplyr::filter(gender == "M", year == 2017) %>%
  .$term_avg
m_nonterm_avg <- workforce_trend%>%
  dplyr::filter(gender == "M", year == 2017) %>%
  .$nonterm_avg
m_rates <- workforce_trend %>%
  dplyr::filter(year == 2017) %>%
  dplyr::mutate(type = "m_rates",
                year1 = total + total * join_avg - total * m_term_avg - total * nonterm_avg) %>%
  dplyr::mutate(year2 = year1 + year1 * join_avg - year1 * m_term_avg - year1 * nonterm_avg) %>%
  dplyr::mutate(year3 = year2 + year2 * join_avg - year2 * m_term_avg - year2 * nonterm_avg) %>%
  dplyr::mutate(year4 = year3 + year3 * join_avg - year3 * m_term_avg - year3 * nonterm_avg) %>%
  dplyr::mutate(year5 = year4 + year4 * join_avg - year4 * m_term_avg - year4 * nonterm_avg) %>%
  dplyr::select(gender, type, `2017` = total, `2018` = year1, 
                `2019` = year2, `2020` = year3, `2021` = year4, `2022` = year5)

workforce_true <- workforce_trend %>%
  dplyr::filter(gender == "F", year < 2017) %>%
  dplyr::mutate(f_rates = total,
                m_rates = total) %>%
  dplyr::select(gender, year, f_rates, m_rates)

pop <- rbind(f_rates, m_rates) %>%
  tidyr::gather(year, value, -(gender:type)) %>%
  tidyr::spread(type, value) %>%
  dplyr::mutate(f_rates = round(f_rates, 0),
                m_rates = round(m_rates, 0),
                year = as.numeric(year)) %>%
  dplyr::filter(gender == "F")  %>%
  rbind(workforce_true, .) %>%
  tidyr::gather(key, value, -(gender:year)) %>%
  dplyr::filter(year > 2010)
  
ggplot(pop, aes(x = year, y = value, group = key, colour = key))+
  geom_line(stat = "identity")
```

This prediction is clearly too linear and static. Instead, I decomposed the rate of termination for men to obtain its trend, and applied that trend to the female rate of termination. Using that rate of termination, I could calculate what the female workforce would've been each year. 

I chose 2010 as the starting year to see what the impact would've been compounded over a few years. 

```{r forecasting, message=FALSE}
library(forecast)

# How many women would you have had now if you had fired them at the same rate as men since 2011?
workforce_freq <- rbind(workforce_trend %>% mutate(qrt = "1"),
                        workforce_trend %>% mutate(qrt = "3")) %>%
  dplyr::filter(year > 2010) %>%
  ungroup()

m_term <- ts(workforce_freq[workforce_freq$gender == "M", ]$term_rate, frequency = 2) %>%
  decompose(type = "multiplicative")
plot(m_term)

f_term <- ts(workforce_freq[workforce_freq$gender == "F", ]$term_rate, frequency = 2) %>%
  decompose(type = "multiplicative")
plot(f_term)

f_f_forecast <- f_term$seasonal * f_term$random * f_term$trend
f_m_forecast <- f_term$seasonal * f_term$random * m_term$trend

plot(f_f_forecast)
plot(f_m_forecast)

# Calculate the rate of change if 
library(data.table)
workforce_freq_new <- workforce_freq %>%
  dplyr::filter(gender == "F") %>%
  dplyr::mutate(f_m_forecast = c(as.vector(f_m_forecast[2:length(f_m_forecast)]), 2),
                workforce = total) %>%
  dplyr::mutate(forecast_term = total * f_m_forecast) %>%
  dplyr::mutate(prev_term = lag(forecast_term, 1)) %>%
  dplyr::mutate(change = new - resigned - retired - prev_term) %>%
  dplyr::select(year, gender, workforce, change) %>%
  setDT(.)

# Calculate what the workforce would've been at this rate of change each year
for (row in 3:nrow(workforce_freq_new)){
  new_workforce <- workforce_freq_new[row - 1L, workforce + change]
  workforce_freq_new[row, workforce := new_workforce]
}
workforce_freq_new <- workforce_freq_new[, .(forecast_workforce = mean(workforce)), by = "year"] %>%
  rbind(., workforce_trend %>% 
          filter(year <= 2010 & gender == "F") %>% 
          ungroup () %>% select(year, forecast_workforce = total))


# Combine the actual female workforce and the forecasted workforce
workforce_results <- workforce_freq %>%
  dplyr::select(-qrt) %>%
  unique(.) %>%
  dplyr::filter(gender == "F") %>%
  dplyr::select(year, total) %>%
  rbind(., workforce_trend %>% 
          filter(year <= 2010 & gender == "F") %>% 
          ungroup () %>% 
          select(year, total)) %>%
  merge(., workforce_freq_new, by = "year") %>%
  dplyr::select(year, 
                "Actual" = total, 
                "Forecast" = forecast_workforce) %>%
  tidyr::gather(key, value, -year)

# Plot
ggplot(workforce_results[workforce_results$year > 2005, ], 
       aes(x = year, y = value, colour = reorder(key, desc(key))))+
  geom_line()+
  geom_line(data = workforce_results[workforce_results$year <= 2010 & 
                                       workforce_results$year > 2005, ])+
  # geom_smooth(data = workforce_results[workforce_results$year > 2009, ], 
  #             se = FALSE)+
  labs(title = "If women had been fired at the same rate as men...",
       subtitle = "By now, you'd have nearly 100 more women",
       x = "Year",
       y = "Number of women in the workforce",
       colour = " ")+
  BIT_theme+
  scale_colour_manual(values = c(strong_palette[2], strong_palette[1]))
```



